Subsetting plant species list by fruiting species

### 2024-06-10 code/L1/L1_P3_TropicalAndes_plant_fruiting_subset.Rmd
update to create species list with any species that has information related to a fruiting trait

---
title: "Subsetting plant species list by fruiting species"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant occurrence and trait data by fruiting plant species."
data input: "ADD"
data output: "ADD"
date: "2024-05-29"
output: html_document
---

@@

# Set file paths
```{r}
data_path_L0<-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
```

@@

```{r}
traits <- rbind(TRY_filtered_subset, BIEN_filtered_subset, GIFT_filtered_subset)
nrow(traits)
```

@@

```{r}
traits %>%
  count(TraitName)
```

```{r}
traits %>%
  distinct(Accepted_species, TraitName) %>%
  count(TraitName)
```

```{r}
# Group by TraitName and summarize the unique TraitValues
unique_trait_values <- traits %>%
  group_by(TraitName) %>%
  summarize(UniqueValues = list(unique(TraitValue)))%>%
  unnest(UniqueValues)
```

# Function to get one value per species
```{r}
# categorical data
factor_data_merge <- function(data, traitname) {
# remove duplicate rows
data <- data %>%
  distinct(.keep_all = TRUE)
# remove rows with missing or empty species names
data <- data[!is.na(data$Accepted_species) & data$Accepted_species != "", ]
# Subset data to just species and TraitValue
data <- data[,c("Accepted_species", "TraitValue")]
# Find the most common trait value for each species
  most_common_values <- aggregate(. ~ Accepted_species, data, function(x) {
    # Find the mode (most common value) of the trait values
    mode_value <- names(sort(table(x), decreasing = TRUE))[1]
    return(mode_value)
  })
# Merge most common trait values back to the original data
merged_data <- merge(data, most_common_values, by = "Accepted_species", all.x = TRUE)

# Rename the merged column
colnames(merged_data)[ncol(merged_data)] <- "TraitValue"

# Remove column TraitValue.x
merged_data <- merged_data[, !colnames(merged_data) %in% "TraitValue.x", drop = FALSE]

# remove duplicate rows
merged_data <- merged_data %>%
  distinct(.keep_all = TRUE)

# Add unit & trait column
merged_data$TraitName <- traitname

  return(merged_data)
}
```

# Fruit type
```{r}
traits_fruitType <- traits %>%
  filter(grepl('Fruit type|fruit type|fruit_type', TraitName))
traits_fruitType$OrigTraitValue <- traits_fruitType$TraitValue

nrow(traits_fruitType)
length(unique(traits_fruitType$Accepted_species))
```

```{r}
traits_fruitType <- traits_fruitType %>%
  filter(is.na(as.numeric(TraitValue)))

nrow(traits_fruitType)
length(unique(traits_fruitType$Accepted_species))
```

```{r}
traits_fruitType %>%
  count(TraitValue)
```

```{r}
# standardize trait values
traits_fruitType <- traits_fruitType %>%
  filter() %>%
  mutate(TraitValue= case_when(
    grepl('Berry|berry|Berry ', OrigTraitValue, ignore.case = TRUE) ~ "berry",
    grepl('capsule|Capsule|Capsule ', OrigTraitValue, ignore.case = TRUE) ~ "capsule",
    grepl('follicle|Follicle', OrigTraitValue, ignore.case = TRUE) ~ "follicle",
  )) %>%
  filter(!is.na(TraitValue))
```

```{r}
# standardize trait name
traits_fruitType <- traits_fruitType %>%
  mutate(TraitName = "FruitType")
```

```{r}
averagetraits_fruitType <- factor_data_merge(traits_fruitType, "Fruit Type")
```

```{r}
nrow(averagetraits_fruitType)
length(unique(averagetraits_fruitType$Accepted_species))
```

```{r}
summary(averagetraits_fruitType)
```

>>>>

---
title: "Subsetting plant species list by fruiting species"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant occurrence and trait data by fruiting plant species."
data input: "TropicalAndes_GBIF_plant_occ_harmonized.csv, TropicalAndes_TRY_plant_traits_harmonized.csv,
             TropicalAndes_BIEN_plant_traits_harmonized.csv, TropicalAndes_GIFT_plant_traits_harmonized.csv"
data output: "TropicalAndes_GBIF_plant_occ_harmonized_subset.csv, TropicalAndes_all_plant_traits_harmonized_subset.csv"
date: "2024-05-29"
output: html_document
---

@@

# Set file paths
```{r}
data_path_L0<-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L0')
data_path_L1 <-file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
output_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/L1')
figure_path <- file.path('G:/Shared drives/SpaCE_Lab_FRUGIVORIA/data/plants/figures')
```

@@

```{r}
# add trait database source column to each subset
TRY_filtered_subset$DatabaseSource <- "TRY"
BIEN_filtered_subset$DatabaseSource <- "BIEN"
GIFT_filtered_subset$DatabaseSource <- "GIFT"

# combine all subsets into one dataframe
traits <- rbind(TRY_filtered_subset, BIEN_filtered_subset, GIFT_filtered_subset)
dim(traits)
```

@@

```{r}
traits_clean %>%
  count(TraitName)
```

```{r}
traits_clean %>%
  distinct(Accepted_species, TraitName) %>%
  count(TraitName)
```

# Create a species list of species with fruiting traits
```{r}
fruit_traits <- c("Fruit dehiscence type", "Fruit dry mass", "Fruit length", "Fruit surface type",
                  "Fruit texture inside (fleshy, leathrey, …)", "Fruit type", "fruit type", "fruit_dryness", "fruit_length_max",
                  "fruit_length_mean", "fruit_length_min", "fruit_type", "maximum fruit length", "minimum fruit length")

# Filter the dataframe for fruiting traits
fruiting_df <- traits_clean %>%
  filter(TraitName %in% fruit_traits)
dim(fruiting_df)

# Extract unique species names
fruiting_species <- unique(fruiting_df$Accepted_species)
length(fruiting_species)
```

# Subset trait data and occurrence data by fruiting species list

```{r}
dim(plant_occ)
plant_occ_subset <- plant_occ %>%
  filter(Accepted_species %in% fruiting_species)
dim(plant_occ_subset)
```

```{r}
dim(traits_clean)
traits_subset <- traits_clean%>%
  filter(Accepted_species %in% fruiting_species)
dim(traits_subset)
```

```{r}
traits_subset %>%
  distinct(Accepted_species, TraitName) %>%
  count(TraitName)
```

# Write data to csv
```{r}
write.csv(plant_occ_subset, file.path(output_path, "TropicalAndes_GBIF_plant_occ_harmonized_subset.csv"))
write.csv(traits_subset, file.path(output_path, "TropicalAndes_all_plant_traits_harmonized_subset.csv"))
```
